---
title: "Model selection in multilevel models"
subtitle: "A practical tutorial in R"
author: "Danaia Burtseva"
date: "February 2026"
output:
  xaringan::moon_reader:
    lib_dir: libs
    seal: false
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE}
options(width = 100)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.width=8, fig.height=4, fig.retina=2)
library(lme4)
library(lmerTest)
library(dplyr)
library(ggplot2)
library(performance)
library(MuMIn)
library(cAIC4)
```

## Learning goals
- What “model selection” means in multilevel (mixed-effects) models
- Compare tools: LRT, AIC/BIC, cAIC, R², model averaging
- Apply to clustered data + crossed random effects

---
## Why it’s harder than regression
- You choose fixed effects *and* random-effects structure
- Variance parameters can be on boundaries (e.g., variance = 0)
- Different criteria answer different questions (prediction vs inference)

---
class: inverse, center, middle
# Part 1 — Clustered data (random intercepts & slopes)

---
## Simulate clustered data
```{r}
set.seed(1)
m  <- 40; ni <- 30; N <- m*ni
group <- factor(rep(seq_len(m), each = ni))
x <- rnorm(N)
b0 <- rnorm(m, 0, 1.0)
b1 <- rnorm(m, 0, 0.6)
eta <- 1.5 + 0.8*x + b0[group] + b1[group]*x
y   <- eta + rnorm(N, 0, 1.2)
dat <- data.frame(y, x, group)
head(dat)
```

---
## Candidate models
```{r}
m0 <- lmer(y ~ 1 + (1|group), data = dat, REML = TRUE)
m1 <- lmer(y ~ x + (1|group), data = dat, REML = TRUE)
m2 <- lmer(y ~ x + (1 + x|group), data = dat, REML = TRUE)
m3 <- lmer(y ~ x + I(x^2) + (1 + x|group), data = dat, REML = TRUE)
```

---
## LRT for fixed effects: use ML (not REML)
```{r}
m1_ml <- update(m1, REML = FALSE)
m3_ml <- update(m3, REML = FALSE)
anova(m1_ml, m3_ml)
```

---
## AIC/BIC across candidates (ML fits)
```{r}
mods_ml <- list(
  m0 = update(m0, REML = FALSE),
  m1 = update(m1, REML = FALSE),
  m2 = update(m2, REML = FALSE),
  m3 = update(m3, REML = FALSE)
)
ic <- do.call(AIC, mods_ml) |> as.data.frame()
ic$BIC <- sapply(mods_ml, BIC)
ic[order(ic$AIC), ]
```

---
## cAIC (conditional AIC)
```{r}
cAIC_vals <- sapply(list(m0=m0, m1=m1, m2=m2, m3=m3), function(mod) cAIC4::cAIC(mod)$caic)
cAIC_vals
```

---
## R² (marginal vs conditional)
```{r}
performance::r2_nakagawa(m2)
```

---
class: inverse, center, middle
# Part 2 — Crossed random effects (subjects × items)

---
## Simulate crossed design
```{r}
set.seed(2)
n_subj <- 60; n_item <- 40
subj <- factor(rep(seq_len(n_subj), each = n_item))
item <- factor(rep(seq_len(n_item), times = n_subj))
cond <- rep(rep(c(0,1), each = n_item/2), times = n_subj) |> factor()
Xcond <- ifelse(cond=="1", 0.5, -0.5)
u_subj0 <- rnorm(n_subj, 0, 0.8); u_item0 <- rnorm(n_item, 0, 0.6)
u_subj1 <- rnorm(n_subj, 0, 0.4); u_item1 <- rnorm(n_item, 0, 0.2)
eta <- 300 + 12*Xcond + u_subj0[subj] + u_item0[item] + u_subj1[subj]*Xcond + u_item1[item]*Xcond
rt  <- eta + rnorm(length(eta), 0, 25)
dat2 <- data.frame(rt, cond, subj, item)
```

---
## Candidate random structures (ML)
```{r}
r0  <- lmer(rt ~ cond + (1|subj) + (1|item), data = dat2, REML = FALSE)
rS  <- lmer(rt ~ cond + (1+cond|subj) + (1|item), data = dat2, REML = FALSE)
rI  <- lmer(rt ~ cond + (1|subj) + (1+cond|item), data = dat2, REML = FALSE)
rSI <- lmer(rt ~ cond + (1+cond|subj) + (1+cond|item), data = dat2, REML = FALSE)
AIC(r0, rS, rI, rSI) |> as.data.frame() |> arrange(AIC)
```

---
## LRT for nested random structures
```{r}
anova(r0, rS)
anova(r0, rI)
anova(rS, rSI)
```

---
## Model averaging (Akaike weights)
```{r}
cand <- list(r0=r0, rS=rS, rI=rI, rSI=rSI)
aic_tab <- MuMIn::model.sel(cand)
aic_tab
avg <- MuMIn::model.avg(cand)
summary(avg)
```

---
## Reporting checklist
- Define your candidate set (justify scientifically)
- Use ML for fixed-effect comparisons; be cautious with boundary issues
- Report AIC/BIC/cAIC + key diagnostics + R²
- If several models are similarly supported: consider model averaging


---
## Session info
```{r}
sessionInfo()
```

---
class: center, middle
# End
